---
title: "Descriptive Group-Level Summaries"
author: "<br><br><span style='font-size:25px;'><strong>Lesley Chapman Hannah, Ph.D., M.S.</strong></span><br>College of Graduate Studies<br>Northeast Ohio Medical University"

format: pptx

execute:
  echo: true
  warning: false
  message: false
---

```{r, echo=FALSE}
library(tidyverse)
library(scales)
library(GGally)

theme_set(theme_minimal(base_size = 18))
set.seed(2026)
```


### Lecture Goal(s)

**Central focus:**

- descriptive analysis often compares groups defined by treatment, disease status, or cohort membership

- discuss how group summaries and comparative plots reveal differences in location, spread, and distribution

Here we ask:

> **How do group summaries and comparative plots reveal differences in location, spread, and distribution?**




------------------------------------------------------------------------

### Descriptive group analysis

::: {style="font-size:0.67em"}

Descriptive Group Analysis involves:

- summarizing the full sample 
- stratifying by a clinically meaningful grouping variable 
- visualizing group-specific patterns
- interpreting scientifically meaningful patterns and trends

Where

-   **overall summary**: single description computed across all observations together
-   **group-specific summary**: separate description computed within each subgroup
-   **comparative plot**: visualization designed to compare groups directly
-   **scientific interpretation**: the biomedical meaning assigned to observed differences in center, spread, or shape

:::

----------------------------------------------------------------------


## Lecture Roadmap

::: {style="font-size:0.67em"}


**Key Terms**

-   **group**: a set of observations sharing the same category, such as treatment arm or disease class
-   **location**: where values tend to be centered, often described by the mean or median
-   **spread**: how variable the observations are, often described by the standard deviation or IQR
-   **distribution**: the overall pattern of values, including center, spread, skewness, and tail behavior
-   **heterogeneity**: meaningful variation across subgroups rather than within one pooled population alone

**Tools [select]**:

-   mean and median by group\
-   summary tables\
-   boxplots\
-   violin plots\
-   interpretation of subgroup heterogeneity
:::

---------------------------

### Group Comparisons: Background

- `group comparison` :  asks whether the observed values of a variable look different once the sample is partitioned by a biologically or clinically meaningful categorical variable such as treatment arm, disease status, or cohort membership

- `group-level summaries and comparative plots` : describe how a variable behaves **within user specified subgroups** and then comparing those subgroup-specific descriptions across groups

**General Applications**

-   treatment groups may differ in average biomarker level
-   disease subtypes may differ in variability, not only in average level
-   two groups may have similar means but very different internal distributions
-   descriptive comparison is often the first step before any formal modeling or hypothesis testing


------------------------------------------------------------------------

### Group Comparisons: Background

::: {style="font-size:0.67em"}
- `summary tables` : report either - median by group or means by group 
- `mean`: measure of central tendency defined as the sum of all observed values divided by the number of observations

  - every value contributes to the mean therefore very high or very low observations can pull the mean upward or downward

- `median` : measure of central tendency defined as the middle value of a set of ordered observations 

  - most informative when we want a measure of central location that is less sensitive to skewness or unusually large values
  - right-skewed biomarkers $\rightarrow$ median often aligns more closely with the value observed for a typical observations [i.e.: patient]

- `median by group`: subgroup-specific measure of central tendency defined as the middle value of the ordered observations within each group

:::

------------------------------------------------------------------------

### Approach for Comparing Groups

::: {style="font-size:0.67em"}
**Key things to ask when comparing groups**

-   **sample size**: are the groups similar in size, or are some summaries based on much smaller numbers of observations?
-   **location**: do the means or medians differ across groups?
-   **spread**: do the standard deviations or IQRs differ across groups?
-   **mean-median agreement**: are the mean and median similar within groups, or do they diverge in ways that suggest skewness?
-   **clinical magnitude**: are the observed differences large enough to matter scientifically, not only numerically?
 
:::
------------------------------------------------------------------------

### Plots for comparing groups

::: {style="font-size:0.67em"}

 `boxplot` : graphical summary of a quantitative variable that represents the ordered structure of the data using key percentiles and tail behavior
 
 Display the following:

- minimum (or lower whisker endpoint)
- first quartile (Q1 - 25th percentile)
- median (Q2 - 50th percentile)
- third quartile (Q3 - 75th percentile)
- maximum (or upper whisker endpoint)

`violin plot`: graphical display of a quantitative variable that combines a smoothed estimate of the distribution (density) with key features of the data’s central tendency and spread

  - **width of the violin** → proportional to the density of observations at a given value
  - **central tendency** → location of the median within the distribution
  - **multimodality** → presence of multiple peaks (subpopulations)

**Considerations for plot-based comparison across groups**

-   is one group positioned higher or lower overall on the y-axis?
-   do some groups show stronger skewness or longer tails?
-   do any groups contain more extreme values?
-   does the shape suggest internal subgroup structure or heterogeneity within a treatment arm?

:::

------------------------------------------------------------------------


### Subgroup Heterogeneity

::: {style="font-size:0.67em"}

**subgroup heterogeneity** : 

  - observations within a group are not all tightly clustered around one common pattern
  - presence of meaningful differences in the distribution, central tendency, variability, or relationships within distinct subgroups of a population

- descriptive statistics - subgroup heterogeneity may appear as:

  -   wide IQR
  -   large standard deviation
  -   long whiskers in a boxplot
  -   wide or irregular violin shape
  -   multiple peaks in a histogram or violin plot
  -   large differences between low and high observations within the same group

:::

---------------------------------------------

### Study Context and Dataset Overview

::: {style="font-size:0.67em"}
Here we evaluate data from a randomized immunotherapy study designed to evaluate how treatment intensity influences both immune system activity and clinical outcomes

**Research Question**

> How do we describe and compare distributions across groups in biomedical data?

> How does the distribution differ by treatment or disease category?

Immunotherapy study follows patients in three arms:

- placebo
- low dose immunotherapy
- high dose immunotherapy

For each patient, the study records multiple outcomes that reflect different dimensions of response:

- **interferon-gamma [IFN-γ] concentration**: continuous biomarker of immune activation
- **CD4+ T-cell count**: continuous measure of immune system status
- **clinical response class** [i.e.: responder vs non-responder or ordinal response categories] : categorical outcome reflecting treatment efficacy

:::
------------------------------------------------------------------------

### Simulated Immunotherapy Dataset

::: {style="font-size:0.67em"}
```{r, echo=FALSE}
set.seed(305)

group_df <- tibble(
  patient_id = sprintf("G%03d", 1:210),
  treatment = sample(c("Placebo", "Low dose", "High dose"), 210, replace = TRUE),
  ifng = round(rlnorm(
    210,
    meanlog = c("Placebo" = log(11), "Low dose" = log(15), "High dose" = log(20))[treatment],
    sdlog = 0.42
  ), 2),
  cd4_count = round(rnorm(
    210,
    mean = c("Placebo" = 510, "Low dose" = 575, "High dose" = 645)[treatment],
    sd = 95
  )),
  response_class = sample(c("Responder", "Stable", "Progressive"), 210, replace = TRUE,
                          prob = c(0.35, 0.40, 0.25))
)
```


```{r}
group_df |> slice_head(n = 8)
```


**Column descriptions**

- `patient_id`: unique identifier for each patient in the study  
- `treatment`: treatment arm assigned to the patient [`Placebo`, `Low dose`, `High dose`] 
  - primary grouping variable used for comparisons across subgroups  
- `ifng`: interferon-gamma concentration, a continuous biomarker of immune activation measured for each patient  
- `cd4_count`: CD4 T-cell count - a continuous measure of immune system status for each patient  
- `response_class`: clinical response category for each patient [`Responder`, `Stable`, `Progressive`]
  - categorical outcome describing treatment response
:::
----------------------------------------------------------------------

### IFN-gamma: Mean by Treatment Group

::: {style="font-size:0.67em"}
```{r}
group_df |>
  group_by(treatment) |>
  summarise(
    n = n(),
    mean_ifng = mean(ifng),
    sd_ifng = sd(ifng)
  )
```

- `mean`: summarizes the average level within each treatment arm
- `standard deviation`:  shows variability among participants in the same arm

**Interpretation**

-   `n` : number of patients that contribute to each treatment-specific summary
-   `mean_ifng`: compares average IFN-gamma across arms and essentially comparing group **location**
-   `sd_ifng` compares within-group variability and therefore compares group **spread**
-   If the high-dose arm has the largest mean, that suggests a higher central level of IFN-gamma in that group
-   If one arm has a larger standard deviation, that suggests greater patient-to-patient heterogeneity within that treatment condition


:::
-------------------------------------------------------------------

### IFN-gamma: Mean by Treatment Group
::: {style="font-size:0.67em"}
```{r}
group_df |>
  group_by(treatment) |>
  summarise(
    n = n(),
    mean_ifng = mean(ifng),
    sd_ifng = sd(ifng)
  )
```

**Interpretation**

- `treatment`: treatment arm defining each subgroup (`Placebo`, `Low dose`, `High dose`)  
- `n`: number of patients within each treatment group (sample size for the subgroup)  
- `mean_ifng`: average IFN-gamma concentration within the group  
- `sd_ifng`: standard deviation of IFN-gamma within the group, quantifying variability around the mean  

- IFN-gamma levels increase across treatment intensity:
  - Placebo ≈ 12.6 pg/mL  
  - Low dose ≈ 16.4 pg/mL  
  - High dose ≈ 20.7 pg/mL  

- variability is present in all groups [SD ≈ 7–8] $\rightarrow$ indicating spread around the mean within each treatment arm

- data suggest that IFN-gamma is right-skewed therefore the mean may be influenced by higher values  
- motivates comparing **median and IQR** as an additional summary of central tendency and spread

:::

-------------------------------------------------------------------


### IFN-gamma: Median and IQR by Treatment Group

::: {style="font-size:0.67em"}
```{r}
group_df |>
  group_by(treatment) |>
  summarise(
    median_ifng = median(ifng),
    iqr_ifng = IQR(ifng),
    median_cd4 = median(cd4_count),
    iqr_cd4 = IQR(cd4_count)
  )
```

**Interpretation**

- `treatment`: treatment arm defining each subgroup  
- `median_ifng`: middle IFN-gamma value within each group 
- `iqr_ifng`: spread of the middle 50% of IFN-gamma values within each group  
- `median_cd4`: middle CD4 count within each group  
- `iqr_cd4`: spread of the middle 50% of CD4 values within each group  - `IFN-gamma and CD4`: show increasing central tendency across treatment groups**  
- `IFN-gamma` : shows increasing variability - consistent with a right-skewed biomarker  
- `median and IQR`: describe how the central bulk of the distribution shifts and spreads across groups 
:::
------------------------------------------------------------------------

### IFN-gamma: Median and IQR by Treatment Group

::: {style="font-size:0.67em"}
```{r}
group_df |>
  group_by(treatment) |>
  summarise(
    median_ifng = median(ifng),
    iqr_ifng = IQR(ifng),
    median_cd4 = median(cd4_count),
    iqr_cd4 = IQR(cd4_count)
  )
```

**Interpretation**

- Median and IQR show that both biomarker levels and their central spread increase with treatment intensity, with IFN-gamma exhibiting greater variability at higher doses

**IFN-gamma median increases with treatment intensity**: shift in the typical biomarker level across groups

  - Placebo ≈ 11.4 pg/mL  
  - Low dose ≈ 14.9 pg/mL  
  - High dose ≈ 19.1 pg/mL  

**IQR for IFN-gamma**: greater spread in the central distribution at higher treatment levels  
  - Placebo ≈ 6.68  
  - Low dose ≈ 8.82  
  - High dose ≈ 9.07  

:::

-------------------------------------------------------

### IFN-gamma: Median and IQR by Treatment Group

::: {style="font-size:0.67em"}
```{r}
group_df |>
  group_by(treatment) |>
  summarise(
    median_ifng = median(ifng),
    iqr_ifng = IQR(ifng),
    median_cd4 = median(cd4_count),
    iqr_cd4 = IQR(cd4_count)
  )
```

**Interpretation**

**CD4 median also increases across treatment**: higher typical immune cell counts with increased treatment intensity  

  - Placebo ≈ 538  
  - Low dose ≈ 567  
  - High dose ≈ 647  

**CD4 IQR** :  variability is similar for placebo and low dose, but more spread in the high dose group

  - Placebo ≈ 118  
  - Low dose ≈ 118  
  - High dose ≈ 158  

:::

--------------



### IFN-gamma: group comparisons

::: {style="font-size:0.67em"}

```{r, echo = FALSE}
plot <- ggplot(group_df, aes(x = treatment, y = ifng)) +
  geom_boxplot() +
  labs(
    title = "IFN-gamma by Treatment Group",
    x = "Treatment group",
    y = "IFN-gamma (pg/mL)"
  )

plot
```

**Interpretation**

- plot shows a dose-dependent increase in immune response with moderate variability across patients
- median IFN-gamma increases from ~11.4 [placebo] to ~14.9 [low dose] to ~19.1 pg/mL [high dose]

Median Comparison:

- high dose: median ≈ 19.1 pg/mL
- low dose: median ≈ 14.9 pg/mL
- placebo: median ≈ 11.4 pg/mL


:::
------------------------------------------------------------------------

### IFN-gamma: group comparisons

::: {style="font-size:0.67em"}

```{r, echo = FALSE}
plot <- ggplot(group_df, aes(x = treatment, y = ifng)) +
  geom_boxplot() +
  labs(
    title = "IFN-gamma by Treatment Group",
    x = "Treatment group",
    y = "IFN-gamma (pg/mL)"
  )

plot
```

**Interpretation**

IQR Comparison:

- high dose group has the widest central spread $\rightarrow$ more variability in response across  
- consistent IFN-gamma levels across the placebo group $\rightarrow$ more tightly clustered IQR
- high dose: IQR ≈ 9.07 pg/mL
- low dose: IQR ≈ 8.82 pg/mL
- placebo: IQR ≈ 6.68 pg/mL


:::
---------------------------

### IFN-gamma: group comparisons

::: {style="font-size:0.67em"}

```{r, echo = FALSE}
plot <- ggplot(group_df, aes(x = treatment, y = ifng)) +
  geom_boxplot() +
  labs(
    title = "IFN-gamma by Treatment Group",
    x = "Treatment group",
    y = "IFN-gamma (pg/mL)"
  )

plot
```

**Interpretation**

Tails and outliers:

- all groups have higher values extending into the upper tail regions
- high dose outliers: ~38–43 pg/mL
- low dose outliers: up to ~49 pg/mL
- placebo outliers: up to ~53 pg/mL

:::
---------------------------


### IFN-gamma: group comparisons

::: {style="font-size:0.67em"}

```{r, echo = FALSE}
plot2 <- ggplot(group_df, aes(x = treatment, y = ifng)) +
  geom_violin() +
  labs(
    title = "Violin Plot of IFN-gamma by Treatment Group",
    x = "Treatment group",
    y = "IFN-gamma (pg/mL)"
  )

plot2
```



**Key plot interpretation**

-   `violin plot` : shows both the distribution shape and density of IFN-gamma within each treatment group
- width of the violin reflects the density of observations at that IFN-gamma level
- widest sections indicate where values are most concentrated within each group
- overall shape reveals distribution features such as skewness and multimodality

:::

-----------------------


### IFN-gamma: group comparisons

::: {style="font-size:0.67em"}

```{r, echo = FALSE}
plot2 <- ggplot(group_df, aes(x = treatment, y = ifng)) +
  geom_violin() +
  labs(
    title = "Violin Plot of IFN-gamma by Treatment Group",
    x = "Treatment group",
    y = "IFN-gamma (pg/mL)"
  )

plot2
```



**Key plot interpretation**

-   wider regions indicate values that occur more frequently within a group
-   narrow upper regions indicate fewer patients at high IFN-gamma levels
-   asymmetry in the violin can reveal skewness
-   differences in overall vertical placement still reflect differences in group location

:::

-----------------------

### IFN-gamma: group comparisons

::: {style="font-size:0.67em"}

```{r, echo = FALSE}
plot2 <- ggplot(group_df, aes(x = treatment, y = ifng)) +
  geom_violin() +
  labs(
    title = "Violin Plot of IFN-gamma by Treatment Group",
    x = "Treatment group",
    y = "IFN-gamma (pg/mL)"
  )

plot2
```



**Interpretation**

-   **high dose group**: greatest density around ~18–20 pg/mL with additional density extending into higher values [~30–40 pg/mL

- **low dose group**: centered lower with highest density around ~12–15 pg/mL

- **placebo group**: most concentrated around ~9–12 pg/mL

:::

-----------------------

### CD4+ T-cell Count: Group Comparison

::: {style="font-size:0.67em"}
```{r, echo=FALSE}
plot <- ggplot(group_df, aes(x = treatment, y = cd4_count)) +
  geom_boxplot() +
  labs(
    title = "CD4 Count by Treatment Group",
    x = "Treatment group",
    y = "CD4 count"
  )

plot
```


**Analysis Goals**

-   compare median CD4+ T-cell count across treatment groups to assess differences in location
-   compare box heights to assess differences in central spread
-   compare whisker symmetry to assess whether the distributions appear roughly symmetric or more skewed

:::

----------------------------------------------------------

### CD4+ T-cell Count: Group Comparison

::: {style="font-size:0.67em"}
```{r, echo=FALSE}
plot <- ggplot(group_df, aes(x = treatment, y = cd4_count)) +
  geom_boxplot() +
  labs(
    title = "CD4 Count by Treatment Group",
    x = "Treatment group",
    y = "CD4 count"
  )

plot
```


**Interpretation**

-  box plots of CD4+ T-cell count shows a dose-dependent improvement in immune function with moderate variability across patients
- median CD4 count increases from ~538 cells/µL [placebo] to ~567 cells/µL [low dose] to ~647 cells/µL [high dose]
:::

----------------------------------------------------------

### CD4+ T-cell Count: Group Comparison

::: {style="font-size:0.67em"}
```{r, echo=FALSE}
plot <- ggplot(group_df, aes(x = ifng)) +
  geom_histogram(bins = 18) +
  facet_wrap(~ treatment, nrow = 1) +
  labs(
    title = "IFN-gamma Distribution Within Each Treatment Group",
    x = "IFN-gamma (pg/mL)",
    y = "Number of patients"
  )

plot
```


**Interpretation**

-   horizontal position of the bars shows the IFN-gamma range for each treatment arm
-   bar heights show how many patients fall in each interval
-   histograms show that IFN-gamma distributions shift upward from ~11.4 [placebo] to ~14.9 [low dose] to ~19.1 pg/mL [high dose] 
- right-skewness and outliers up to ~50 pg/mL indicates variable but dose-dependent immune response

:::
------------------------------------------------------------------------

### Summary

::: {style="font-size:0.67em"}


-   group-level summaries and plots transform raw biomedical data into interpretable evidence about how biological or clinical processes differ across populations
- computed group-specific summaries:
    - mean → average level [sensitive to extreme values]
    - median → typical value [robust to skewness]
    - standard deviation / IQR → variability within groups
- used comparative visualizations:
    - boxplots → summarize quartiles, spread, and outliers
    - violin plots → reveal full distribution shape and density
- interpreted differences across groups in:
    - location [center of the distribution]
    - spread [variability]
    - distribution shape [skewness, tails, multimodality]

:::
------------------------------------------------------------------------

